Case study validations of automatic bullet matching\tnoteref{t1}\tnoteref{t2}

### 2019-11-10 index.Rnw
match strength -> similarity

\begin{abstract}
Recent advances in microscopy have made it possible to collect 3D topographic data, enabling virtual comparisons based on the collected 3D data as a supplement to traditional comparison microscopy and 2D photography. Automatic matching algorithms have been introduced for various scenarios, such as matching cartridge cases \citep{tai2018fully,song3DTopographyMeasurements2014} or matching bullet striae \citep{aoas2, chuAutomaticIdentificationBullet2013, dekinderAutomatedComparisonsBullet1999}. One key aspect of validating automatic matching algorithms is to evaluate the performance of the algorithm on external tests, that is, using data which were not used to train the algorithm. Here, we present a discussion of the performance of the matching algorithm \citep{aoas2} in three studies conducted using different Ruger weapons. We consider the performance of three scoring measures: random forest score, cross correlation, and consecutive matching striae (CMS) at the land-to-land level and, using Sequential Average Maxima scores, also at the bullet-to bullet level. Cross correlation and random forest scores both result in perfect discrimination of same-source and different-source bullets. At the land-to-land level, discrimination for both cross correlation and random forest scores (based on area under the curve, AUC) is excellent ($\ge$ 0.90).
\end{abstract}

@@

Once signatures have been extracted from stable regions with expressed striae, they must be aligned in order to assess their match strength. Maximized cross correlation is used to pair-wise align signatures \citep{aoas2,vorburgerApplicationsCrosscorrelationFunctions2011}.

@@

Using the lag determined for aligning two signatures, other quantitative features describing the similarity of the two signatures can be extracted. Numerical features such as cross correlation and Euclidean distance
can be computed from the aligned signatures alone. Other features include characteristics of the signature that firearms and toolmark examiners base their visual assessment on, such as striae depth, number of matching striae, and the number of consecutively matching striae (CMS), one of the measures firearms examiners use to describe the strength of a match quantitatively.

In order to evaluate CMS, we must first identify peaks and valleys in each of the signatures, then determine whether these peaks and valleys overlap sufficiently. Features which firearms examiners assess visually depend on the identification of these extrema. If there are at least six consecutive matching striae, the bullets are considered to be a good match in practice \citep{biasotti_firearms_1997,nichols_scientific_2006}. CMS in this setting are counted as the number of consecutively matching peaks in signatures of two aligned lands. Peaks -- rather than peaks and valleys -- are used for this determination in part because it is difficult to visually assess valleys except as relative to peaks. Note that with automatic alignment based on the maximum cross correlation rather than matching striation marks, it is possible for two lands to have a consecutive matching striae of zero.

@@

Cross correlation can also be used to assess match strength between two aligned signatures. Cross correlation varies between -1 and 1, but as the alignment is based on the maximized cross correlation value, in practice the cross correlations for aligned signatures are generally positive. Cross correlation values which are close to 1 indicate similarity between signatures and with it may indicate that the two signatures come from the same source. Low cross correlation values indicate signatures which may originate from different sources.

The random forest presented in \citet{aoas2} is based on a combination of multiple characteristics, including cross correlation, number of matching striae, and number of consecutively matching striae. The output of the random forest is a score between 0 and 1 representing the algorithm's assessment of the strength of the match, where scores close to 1 indicate it to be highly likely that the aligned signatures come from the same source, while scores close to 0 indicate that the aligned signatures are highly likely to originate from different sources.

@@

\subsection{Assessing Bullet-to-Bullet Match Strength}

@@

When imaging bullets, operators scan one land at a time in a clockwise (left twisted rifling) or anti-clockwise (right twisted rifling) sequence. The order in which scans are acquired is kept as meta-information. Let us assume that lands on a bullet are labelled $\ell_i$ with $i = 1, .., p$, where $p$ indicates the number of lands a bullet has, as determined by the rifling of a barrel. For all of the bullets considered here the number of lands $p$ is 6. A match between  a pair of lands from two bullets therefore results in an expected additional $p-1$ matches between pairs of lands. These lands are also expected to be in a sequence, i.e. if there is a match between lands $\ell_i$ on bullet 1 and $\ell_j$ on bullet 2, we also expect lands $\ell_{i\oplus s}$ and $\ell_{j\oplus s}$ to match for all integers $s$, where $\oplus$ is defined as $a \oplus b \equiv \left((a + b - 1)\mod p\right) + 1$. This relationship gives rise to the sequence average maximum (SAM) to quantify a bullet-to-bullet match.

@@

In this paper, we validate the algorithm described in \citet{aoas2} on three external test sets. We assess the results on each set and evaluate whether the minimal requirements described above are fulfilled for bullet-to-bullet SAM scores. We compare the random forest algorithm's performance to the performance of other suggested measures for quantitative assessment of match strength, such as cross correlation and consecutive matching striae.

@@

\section{Evaluating the Random Forest Algorithm}
The random forest SAM scores correctly separate known matches from known non-matches, using SAM scores to aggregate land-to-land scores into a bullet-to-bullet comparison. With no errors at the bullet-to-bullet level, we can now evaluate the scores in light of (R1) and (R2). In this section, we will compare the random forest scores to cross correlation scores; cross correlation is one of the components of the random forest, but has also been used to quantify the strength of a match in its own right \citep{chuPilotStudyAutomated2010}. In addition, both scoring methods use the same scale (0 to 1) and are continuous along that interval; other evaluation methods, such as consecutive matching striae are discrete and more difficult to directly compare to continuous measurements.

@@

Examining land-to-land scores provides more details about the matching algorithm's performance. \autoref{fig:compare-land-to-land} shows density curves for each study. There is some overlap between known matching and known non-matching land-to-land scores; in many cases the random forest scores for these overlapping values are more extreme than the corresponding cross correlation scores. This suggests that the matching algorithm is sensitive to the presence of LEAs which have low land-to-land match scores, such as the matches of bullets 'I' and 'F' in the Hamby 44 study. Comparing this to the bullet-to-bullet score density plot shown in \autoref{fig:compare}, we see that there is much greater separation between the same-source and different-source densities for each study in the bullet-to-bullet comparisons than in the land-to-land comparisons, that is, one or two poorly matching lands does not prevent the bullet from showing a matching score, though if there are weak matches on several lands, such as the marked Hamby 44 comparisons in \autoref{fig:compare}, the overall score may be affected.

@@

At the beginning of this study, we anticipated Hamby 44 would be the easiest set to work with because of its similarity to the sets used to train the random forest algorithm. In a surprise turn of events, it was the hardest to solve, in part because of damage to the bullets that obscured LEA striae. It has been shown \citet{lpr2} that parts of lands can be used for successful identifications. The algorithm proposed by \citet{aoas2} is not capable of  automatically detecting parts of lands with well-expressed striae. It may be useful to couple the \citet{aoas2} algorithm with an algorithm which assesses the quality of the input data and determines which portions of the data to use for comparison. This would emulate the process used by examiners, who first assess the quality of the evidence and whether there is enough information present to attempt a match. One major disadvantage of an automated algorithm is that all of the decisions humans make (recognizing degraded land areas, excluding those areas from consideration, matching only the remaining areas) must be explicitly characterized; however, this explicit characterization means that the process can be scientifically validated to a much higher degree than the human perceptual process.

All of the studies presented here involve the same type of firearm. We see some variations in scores across different models of Rugers, and it is probable that we will see even bigger variations with different brands of firearms that vary in number of lands and land length. We also know that different firearms and ammunition combinations mark differently well \citep{boltonkingPreventingMiscarriagesJustice2016, bonfantiInfluenc1999}. Rugers are among the firearms that mark very well, so we would correspondingly expect to see lowered similarity scores for other firearms. The matching algorithm works on matching lands. Barrels with polygonal rifling, such as Glocks, do not introduce well defined lands on bullets, which makes an application of the matching algorithm impossible.

>>>>

\begin{abstract}
Recent advances in microscopy have made it possible to collect 3D topographic data, enabling virtual comparisons based on the collected 3D data as a supplement to traditional comparison microscopy and 2D photography. Automatic comparison algorithms have been introduced for various scenarios, such as matching cartridge cases \citep{tai2018fully,song3DTopographyMeasurements2014} or matching bullet striae \citep{aoas2, chuAutomaticIdentificationBullet2013, dekinderAutomatedComparisonsBullet1999}. One key aspect of validating these automatic comparison algorithms is to evaluate the performance of the algorithm on external tests, that is, using data which were not used to train the algorithm. Here, we present a discussion of the performance of the matching algorithm \citep{aoas2} in three studies conducted using different Ruger weapons. We consider the performance of three scoring measures: random forest score, cross correlation, and consecutive matching striae (CMS) at the land-to-land level and, using Sequential Average Maxima scores, also at the bullet-to bullet level. Cross correlation and random forest scores both result in perfect discrimination of same-source and different-source bullets. At the land-to-land level, discrimination for both cross correlation and random forest scores (based on area under the curve, AUC) is excellent ($\ge$ 0.90).
\end{abstract}

@@

Once signatures have been extracted from stable regions with expressed striae, they must be aligned in order to assess their similarity, just as examiners would manually align two bullets under a comparison microscope. Maximized cross correlation is used to pair-wise align signatures \citep{aoas2,vorburgerApplicationsCrosscorrelationFunctions2011}.

@@

Using the lag determined for aligning two signatures, other quantitative features describing the similarity of the two signatures can be extracted. Numerical features such as cross correlation and Euclidean distance
can be computed from the aligned signatures alone. Other features include characteristics of the signature that firearms and toolmark examiners base their visual assessment on, such as striae depth, number of matching striae, and the number of consecutively matching striae (CMS), one of the measures firearms examiners use to describe the similarity of two bullets quantitatively.

In order to evaluate CMS, we must first identify peaks and valleys in each of the signatures, then determine whether these peaks and valleys overlap sufficiently. Features which firearms examiners assess visually depend on the identification of these extrema. If there are at least six consecutive matching striae, the bullets are considered to be similar in practice \citep{biasotti_firearms_1997,nichols_scientific_2006}. CMS in this setting are counted as the number of consecutively matching peaks in signatures of two aligned lands. Peaks -- rather than peaks and valleys -- are used for this determination in part because it is difficult to visually assess valleys except as relative to peaks. Utilizing automatic alignment based on maximum cross correlation (as opposed to matching striation marks) allows the possibility that two lands will have zero consecutive matching striae; this would not occur with manual alignment based on striation marks.

@@

Cross correlation can also be used to assess similarity between two aligned signatures. Cross correlation varies between -1 and 1, but as the alignment is based on the maximized cross correlation value, in practice the cross correlations for aligned signatures are generally positive. Cross correlation values which are close to 1 indicate similarity between signatures, and may consequentially indicate that the two signatures come from the same source. Low cross correlation values indicate signatures which may originate from different sources.

The random forest presented in \citet{aoas2} is based on a combination of multiple characteristics, including cross correlation, number of matching striae, and number of consecutively matching striae. The output of the random forest is a score between 0 and 1 representing the algorithm's assessment of the similarity of the two signatures, where scores close to 1 indicate it to be highly likely that the aligned signatures come from the same source, while scores close to 0 indicate that the aligned signatures are highly likely to originate from different sources.

@@

\subsection{Assessing Bullet-to-Bullet Similarity}

@@

When imaging bullets using a confocal light microscope, operators scan one land at a time in a clockwise (left twisted rifling) or anti-clockwise (right twisted rifling) sequence (Other imaging systems, such as BulletTrax 3D\footnote{https://www.ultra-forensictechnology.com/en/our-products/ballistic-identification/bullettrax/} or BalScan\footnote{https://www.forensic.cz/en/products/balscan}, may capture lower-resolution scans of an entire bullet at one time). The order in which scans are acquired is kept as meta-information. Let us assume that lands on a bullet are labelled $\ell_i$ with $i = 1, .., p$, where $p$ indicates the number of lands a bullet has, as determined by the rifling of a barrel. For all of the bullets considered here the number of lands $p$ is 6. A match between  a pair of lands from two bullets therefore results in an expected additional $p-1$ matches between pairs of lands. These lands are also expected to be in a sequence, i.e. if there is a match between lands $\ell_i$ on bullet 1 and $\ell_j$ on bullet 2, we also expect lands $\ell_{i\oplus s}$ and $\ell_{j\oplus s}$ to match for all integers $s$, where $\oplus$ is defined as $a \oplus b \equiv \left((a + b - 1)\mod p\right) + 1$. This relationship gives rise to the sequence average maximum (SAM) to quantify a bullet-to-bullet match.

@@

In this paper, we validate the algorithm described in \citet{aoas2} on three external test sets. We assess the results on each set and evaluate whether the minimal requirements described above are fulfilled for bullet-to-bullet SAM scores. We compare the random forest algorithm's performance to the performance of other suggested measures for quantitative assessment of bullet similarity, such as cross correlation and consecutive matching striae.

@@

\section{Evaluating the Random Forest Algorithm}
The random forest SAM scores correctly separate known matches from known non-matches, using SAM scores to aggregate land-to-land scores into a bullet-to-bullet comparison. With no errors at the bullet-to-bullet level, we can now evaluate the scores in light of (R1) and (R2). In this section, we will compare the random forest scores to cross correlation scores; cross correlation is one of the components of the random forest, but has also been used to quantify the strength of a match in its own right \citep{chuPilotStudyAutomated2010}. In addition, both scoring methods use the same scale (0 to 1) and are continuous along that interval; other evaluation methods, such as consecutive matching striae, are discrete and more difficult to directly compare to continuous measurements.

@@

Examining land-to-land scores provides more details about the matching algorithm's performance. \autoref{fig:compare-land-to-land} shows density curves for each study. There is some overlap between known matching and known non-matching land-to-land scores; in many cases the random forest scores for these overlapping values are more extreme than the corresponding cross correlation scores. This suggests that the matching algorithm is sensitive to the presence of LEAs which have low land-to-land similarity scores, such as the matches of bullets 'I' and 'F' in the Hamby 44 study. Comparing this to the bullet-to-bullet score density plot shown in \autoref{fig:compare}, we see that there is much greater separation between the same-source and different-source densities for each study in the bullet-to-bullet comparisons than in the land-to-land comparisons, that is, one or two poorly matching lands does not prevent the bullet from showing a matching score, though if there are weak matches on several lands, such as the marked Hamby 44 comparisons in \autoref{fig:compare}, the overall score may be affected.

@@

At the beginning of this study, we anticipated Hamby 44 would be the easiest set to work with because of its similarity to the sets used to train the random forest algorithm. In a surprise turn of events, it was the hardest to solve, in part because of damage to the bullets that obscured LEA striae. It has been shown \citet{lpr2} that parts of lands can be used for successful identifications. The algorithm proposed by \citet{aoas2} is not capable of  automatically detecting parts of lands with well-expressed striae. It may be useful to couple the \citet{aoas2} algorithm with an algorithm which assesses the quality of the input data and determines which portions of the data to use for comparison. This would emulate the process used by examiners, who first assess the quality of the evidence and whether there is enough information present to attempt a comparison. One major disadvantage of an automated algorithm is that all of the decisions humans make (recognizing degraded land areas, excluding those areas from consideration, matching only the remaining areas) must be explicitly characterized; however, this explicit characterization means that the process can be scientifically validated to a much higher degree than the human perceptual process.

All of the studies presented here involve the same type of firearm. We see some variations in scores across different models of Rugers, and it is probable that we will see even bigger variations with different brands of firearms that vary in number of lands and land length. We also know that different firearms and ammunition combinations mark differently well \citep{boltonkingPreventingMiscarriagesJustice2016, bonfantiInfluenc1999}. Rugers are among the firearms that mark very well, so we would correspondingly expect to see lowered similarity scores for other firearms. The matching algorithm works on matching lands, so barrels with polygonal rifling, such as Glocks, which do not introduce well defined lands on bullets, cannot be compared using the matching algorithm examined here.
